---
title: "Multivariate Statistical Analysis"
author: Lucas Fellmeth, Helen Kafka, Sven Bergmann
date: 02/07/24
output: pdf_document
---

# Exercise 1
## Problem
Determine which of the following matrices admit a matrix square root, and for the ones that do, find it.

### (i)
```{r}
A1 <- matrix(c(2, 3, -1, 3, 2, 1, -1, 1, 0), 3, 3)
A1
```
Check if the eigenvalues of A1 are positive:

```{r}
e1 <- eigen(A1)$values
e1
```
Because one of the eigenvalues is negative, more precisely -2, according to the lecture it does not exist a square root matrix for A1

### (i)
```{r}
A2 <- matrix(c(3, 2, - 1, 2, 3, 1, -1, 1, 2), 3, 3)
A2
```
Check if the eigenvalues of A2 are positive:
```{r}
e2 <- eigen(A2)$values
e2
```
Because the eigenvalues of A2 are positive and the matrix is symmetric, A2 is positive definite and it has a matrix square root given by the following matrix:

```{r}
L <- diag(eigen(A2)$values)
V <- eigen(A2)$vectors
L_sq <- sqrt(L)
B <- V %*% L_sq %*% solve(V)
B
B %*% B
```
# Exercise 2
## Problem
Let $H$ be the subspace of $\mathbb{R}^3$ with basis $v_1$ and $v_2$ given by
```{r}
v1 <- c(1, 1, 1)
v2 <- c(1, -1, 0)
v1
v2
```
### a)
Find the projection matrix P associated with H:
```{r}
A <- matrix(c(v1, v2), 3, 2)
A
A_t <- t(A)
A_i <- solve(A_t %*% A)
P <- A %*% A_i %*% A_t
P
```
### b)

### i)

```{r}
x1 <- c(0, 1, 1)
x1_best <- P %*% x1
x1_best
```


### ii)
```{r}
x2 <- c(2, 0, 1)
x2_best <- P %*% x2
x2_best
x2 %*% A
```

The result tells us $x_2$ is part of the subspace. We can confirm this because $x_2$ is a linear combination of $v_1$ and $v_2$.


### iii)
```{r}
x3 <- c(-1, -1, 2)
x3_best <- P %*% x3
x3_best
v1 %*% x3
v2 %*% x3
```

The result tells us, $x_3$ is perpendicular to $v_1$ and $v_2$.

# Exercise 3
## Problem
Consider the matrix
```{r}
B <- matrix(c(2 / 3, 1 / 3, 1 / 3, 1 / 3, 2 / 3, -1 / 3, 1 / 3, -1 / 3, 2 / 3), 3, 3)
B
```
Is $B$ a projection matrix for some space $H$? If so, find a basis for $H$.

According to lecture, such that a matrix is a projection matrix it has to be symmetric and idempotent.
```{r}
B
t(B)
B %*% B
```
Because the matrices are the same, $B$ is symmetric and idempotent. Hence, $B$ is a projection matrix.

To find a basis for $H$, we have to find the eigenvectors:
```{r}
eigen(B)$vector
```
