---
title: "Multivariate Statistical Analysis"
subtitle: "Final work"
author: Lucas Fellmeth, Helen Kafka, Sven Bergmann
date: 05/09/24
output: pdf_document
---

```{r, echo = F}
library(knitr)
opts_chunk$set(tidy.opts = list(width.cutoff = 60), tidy = TRUE)
```
```{r, warning = F, message = F, error = F}
library(nnet)
library(caret)
```
# Problem 1

Consider the ``vehicle`` dataset.
```{r}
vehicles <- read.csv(file = './Data_csv/vehicle.csv')
vehicles <- subset(vehicles, select = -classdigit)
vehicles$class <- factor(vehicles$class)
```

The goal is to identify 3D objects from 2D images captured by cameras at different angles.
The objects in this case are four types of vehicles
(identified by the variables ``class`` and ``classdigit``),
and the other 18 numerical variables are measurements extracted from these 2D images.
Split the data into training and test sets (80/20).

```{r}
i_test_vehicles <- sample(seq_along(vehicles[, 1]), size = length(vehicles[, 1]) * 0.2)
vehicles_test <- vehicles[i_test_vehicles,]
vehicles_train <- vehicles[-i_test_vehicles,]
```

On the training set compute:

- a multinomial logistic classifier

```{r}
multinomial_logistic_classifier <- glm(formula = class ~ . - class, data = vehicles_train, family = binomial)
```

- a single-hidden-layer neural network classifier (with the number of hidden nodes to be determined)

```{r}
num_hidden_nodes_vehicles <- 20
nnets_vehicles <- vector(mode = "list", length = num_hidden_nodes_vehicles)
for (i in 1:num_hidden_nodes_vehicles) {
  nnets_vehicles[[i]] <- nnet(class ~ . - class, data = vehicles_train, size = i, trace = F)
}
```

On the test data,
find the cross-classification tables and the misclassification rates.

```{r}
pi1_hat <- predict(multinomial_logistic_classifier, type = 'response', newdata = vehicles_test)
gr_hat <- ifelse(pi1_hat > 0.5, 2, 1)
mctable <- table(gr_hat, vehicles_test$class)
mctable
```
```{r}
1 - sum(diag(mctable)) / length(vehicles_test$class)
```

```{r}
cms_vehicles <- vector(mode = 'list', length = num_hidden_nodes_vehicles)
mscrs_vehicles <- list()
for (i in 1:num_hidden_nodes_vehicles) {
  predictions <- factor(predict(nnets_vehicles[[i]], newdata = vehicles_test, type = 'class'),
                        levels = levels(vehicles_test$class))
  cms_vehicles[[i]] <- confusionMatrix(predictions, vehicles_test$class)
  mscrs_vehicles[i] <- 1 - cms_vehicles[[i]]$overall["Accuracy"]
}
```

```{r}
for (i in 1:num_hidden_nodes_vehicles) {
  cat('Misclassification rate for', i, 'hidden nodes:', mscrs_vehicles[[i]], '\n')
}
```

Which of the above methods is better?
Is there any specific type of vehicle that is harder to classify than the others?
