---
title: "Multivariate Statistical Analysis"
subtitle: "Homework 4"
author: Lucas Fellmeth, Helen Kafka, Sven Bergmann
date: 03/13/24
output: pdf_document
---

```{r, echo = F}
library(knitr)
opts_chunk$set(tidy.opts = list(width.cutoff = 60), tidy = TRUE)
```
```{r}
set.seed(42)
```
# Problem 1

Consider the ``bupa_liver_disorder`` data set,
where a number of individuals are classified into two classes according to blood test results
(see full description on p. 258 of the book).

```{r}
bupa_liver_disorder <- read.csv(file = '../Data_csv/bupa_liver_disorder.csv')
bupa_liver_disorder$class <- factor(bupa_liver_disorder$class)
```

## a)

Split the data set into training and test sets (roughly a 70/30 split).
Compute the logistic classifier using the training set.

```{r}
i_test <- sample(seq_along(bupa_liver_disorder[,1]), size = length(bupa_liver_disorder[,1]) * 0.3)
bupa_liver_disorder_test <- bupa_liver_disorder[i_test,]
bupa_liver_disorder_train <- bupa_liver_disorder[-i_test,]
```
```{r}
out <- glm(formula = class ~ mcv + alkphos + sgpt + sgot + gammagt + drinks, data = bupa_liver_disorder_train, family = binomial)
out$coefficients
```

## b)

Construct the misclassification table and compute the misclassification rate for the test set.

```{r}
pi1_hat <- predict(out, type = 'response', newdata = bupa_liver_disorder_test)
gr_hat <- ifelse(pi1_hat > 0.5, 1, 2)
mctable <- table(gr_hat, bupa_liver_disorder_test$class)
mctable
```
```{r}
1 - sum(diag(mctable)) / length(bupa_liver_disorder_test$class)
```

# Problem 2

Consider the ``ecoli`` data set.
These data were obtained in a study of protein localization sites for 336 examples of E. coli.
There are 7 predictor variables and a class variable, localization_site, which indicates the protein localization.

```{r}
ecoli <- read.csv(file='../Data_csv/ecoli.csv')
ecoli$localization_site <- factor(ecoli$localization_site)
```

## a)

Split the data set into training and test sets (roughly a 80/20 split).
Compute the multinomial logistic classifier using the training set.

```{r}
i_test <- sample(seq_along(ecoli[,1]), size = length(ecoli[,1]) * 0.2)
ecoli_test <- ecoli[i_test,]
ecoli_train <- ecoli[-i_test,]
```
```{r}
out <- nnet::multinom(localization_site ~ mvg + gvh + lip + chg + aac + alm1 + alm2,
                data = ecoli,
                maxit = 100)
out
```
```{r}
mctable <- table(predict(out, newdata = ecoli_test), ecoli_test$localization_site)
```
```{r}
1 - sum(diag(mctable)) / length(ecoli_test$localization_site)
```
