---
title: "Multivariate Statistical Analysis"
subtitle: "Homework 7"
author: Lucas Fellmeth, Helen Kafka, Sven Bergmann
date: 04/30/24
output: pdf_document
---

```{r, echo = F}
library(knitr)
opts_chunk$set(tidy.opts = list(width.cutoff = 60), tidy = TRUE)
```
# Problem 1

The ``British_towns`` dataset contains distances between 48 British towns.

```{r}
towns <- read.csv(file = 'Data_csv/British_towns.csv')
```

## a)

Carry out a metric multidimensional scaling.
From a plot of the eigenvalues,
does a two-dimensional solution appear reasonable?


## b)

Plot the two-dimensional solution (as a 2D scatterplot) labeling each point by town name.
Compare with a map of England.
Do they look similar?


# Problem 2

Consider the zoo dataset,
where 17 variables are observed on 101 animals.
Excluding the variable ``type``, the other 16 variables are categorical:
most of them are yes/no indicators,
and ``legs`` can be seen as a categorical variable too.
So, the dissimilarity between animals i and j can be defined as the number of characteristics in which they differ:
$\delta_{ij} = \#\{f : x_{ik} \neq x_{jk}\}$.
In ``R`` you can do this as follows:
``D <- matrix(nrow=101,ncol=101)
for (i in 1:101){ for (j in 1:101){ D[i,j] <- sum(x[i,]!=x[j,]) }}``

```{r}
zoo <- read.csv(file = 'Data_csv/zoo.csv')
```

## a)

Carry out a two-dimensional multidimensional scaling.
Plot the resulting configuration using animal names as labels.

## b)

Try to find, if possible, an interpretation for the coordinates, in terms of anatomical or physiological characteristics of the animals.