---
title: "Multivariate Statistical Analysis"
subtitle: "Homework 6"
author: Lucas Fellmeth, Helen Kafka, Sven Bergmann
date: 04/17/24
output: pdf_document
---

```{r, echo = F}
library(knitr)
opts_chunk$set(tidy.opts = list(width.cutoff = 60), tidy = TRUE)
```
```{r, warning = F, message = F, error = F}

```
```{r}
set.seed(42)
```
# Problem 1

## a)

Carry out a hierarchical agglomerative cluster analysis of the ``primate_scapulae`` data using complete linkage.
Use the variables AD.BD, AD.CD, EA.CD, Dx.CD, SH.ACR, EAD and $\beta$ only.
(Remember to scale the variables prior to clustering.)
Does the dendrogram suggest a definite number of clusters as the best solution?
If not, what would be reasonable solutions?

```{r}
primate_sculpae <- read.csv(file = 'Data_csv/primate_scapulae.csv')
```

## b)

Decide on a reasonable solution in part (a) and assess that solution using a silhouette plot.

## c)

Is there any correspondence between the clusters obtained above
and the primate species as indicated by the variable class?