---
title: "Multivariate Statistical Analysis"
subtitle: "Homework 2"
author: Lucas Fellmeth, Helen Kafka, Sven Bergmann
date: 02/15/24
output: pdf_document
---

```{r, echo = F}
library(knitr)
opts_chunk$set(tidy.opts = list(width.cutoff = 80), tidy = TRUE)
```
```{r, warning = F, message = F}
library(expm)
```

# Problem 1

## (a)

We have $x\in\mathbb{R}^p$, $\mathbb{E}[x] = \mu$, $\text{Cov}(x) = \Sigma$, $A$ is a $p \times p$ constant matrix and $\text{tr}(A v v^\top) = v^\top A v$.
Because $A$ is a constant $p \times p$ matrix, $A$ is symmetric.
Because of this symmetry, it follows that it has a Cholesky decomposition as $A = C^\top C$.

Let $y = C x$.

Then

\begin{align*}
    \mathbb{E}[x^\top A x]
    &= \mathbb{E}[x^\top C^\top C x] \\
    &= \mathbb{E}[(Cx)^\top C x] \\
    &= \mathbb{E}[y^\top y] \\
    &= \sum_i \mathbb{E}[y_i^2] \\
    &= \sum_i \text{Var}(y_i) + \mathbb{E}[y_i]^2 \\
    &= \text{tr}(\Sigma_y) + \mu_y^\top \mu_y
\end{align*}

where $\Sigma_y = \mathbb{E}[(y-\mathbb{E}[y])(y-\mathbb{E}[y])^\top) = C \Sigma C^\top$
and $\mu_y = C_\mu$.

\begin{align*}
    \implies \mathbb{E}[x^\top A x]
    &= \text{tr}(C \Sigma C^\top) + \mu^\top \underbrace{C^\top C}_{=A} \mu \\
    &= \text{tr}(\Sigma C^\top C) + \mu^\top A \mu \\
    &= \text{tr}(\Sigma A) + \mu^\top A \mu
\end{align*}

## (b)
## (c)

# Problem 2
## (a)
## (b)
```{r}
Sigma <- matrix(data = c(1, -2, 0, -2, 5, 0, 0, 0, 2), nrow = 3, ncol = 3)
print(Sigma)
```
```{r}
Sigma_sqrt <- sqrtm(Sigma)
print(Sigma_sqrt)
```

# Problem 3
