---
title: "MthStat 768"
date: February 21, 2024
output: pdf_document
---

```{r, echo = F}
library(knitr)
opts_chunk$set(tidy.opts = list(width.cutoff = 60), tidy = TRUE)
```

# Chapter 7: Principal Component Analysis
```{r}
iris <- read.csv(file = '../Data_csv/Iris.csv') # two points for knitting, one point for running it in the cell
```
```{r}
unique(iris$class)
```
```{r}
iris$class <- factor(iris$class)
levels(iris$class)
```
```{r}
levels(iris$class) <- c('Setosa', 'Versicolor', 'Virginica')
levels(iris$class)
```
```{r}
tapply(X = iris$petal_width, INDEX = iris$class, FUN = mean)
```
```{r}
apply(X = iris[, 1:4], MARGIN = 2, FUN = sd)
```
```{r}
iris[, 1:4] <- scale(x = iris[, 1:4])
apply(X = iris[, 1:4], MARGIN = 2, FUN = sd)
```
```{r}
pca <- princomp(~sepal_length +
  sepal_width +
  petal_length +
  petal_width, data = iris)
pca <- princomp(~. - class, data = iris)
summary(pca)
```

The output of `princomp(..)` is a list with elements.
- `sdev`: standard deviation of the components, $\sqrt{\lambda_k}$.
- `loadings`: eigenvectors of v.
- `scores`: component scores $\xi$.

```{r}
lmb <- pca$sdev^2
lmb
```
```{r}
cumsum(lmb) / sum(lmb)
```

Then q=2 gives a good approximation to the data.
$\xi_i \in \mathbb{R}^2$

```{r}
xi <- pca$scores[, 1:2]
xi
```
```{r}
colnames(xi) <- c('PC1', 'PC2')
```
```{r}
plot(xi[, 1], xi[, 2])
```

To see if the species of Iris are spatially separated, we can plot $\xi_{i1}$ vs $\xi_{i2}$.

```{r}
plot(xi[, 1], xi[, 2], col = iris$class, pch = 16)
```
```{r}
plot(xi[iris$class == 'Setosa', 1], xi[iris$class == 'Setosa', 2], col = 'blue', pch = 16, xlim = c(-3, 4), ylim = c(-3, 3))
points(xi[iris$class == 'Virginica', 1], xi[iris$class == 'Virginica', 2], col = 'red', pch = 16)
points(xi[iris$class == 'Versicolor', 1], xi[iris$class == 'Versicolor', 2], col = 'green', pch = 16)
```

We see that Iris Setosa is well separated from the other species.
Iris Virginica and Versicolor are also separated to some extent, but not as neatly as they are from Setosa.

```{r}
pca$loadings[, 1]
```

The eigenvectors are the loadings.
$\xi_{i1}$ is a contrast between the average of (sepal length, petal length, petal width) and sepal width.

---

```{r}
PAH <- read.csv(file = '../Data_csv/PAH.csv')
```
```{r}
PAH <- PAH[, -(1:10)]
```
```{r}
w <- seq(220, 350, by = 5)
```
```{r}
plot(w, PAH[1,], type = 'l', lwd = 2)
```
```{r}
matplot(w, t(PAH), type = 'l', lwd=2)
```
